Reasonably Programmable Syntax

### 2017-05-24 popl18/syntax-popl18.tex
updated abstract

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programmers, 
like pen-and-paper mathematicians, often want to define new ``syntactic sugar'' to  lower the syntactic cost of common idioms. 
Unfortunately, the most direct mechanisms of syntactic control available to programmers, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about syntactic determinism -- separately defined syntax extensions can conflict syntactically with one another. Moreover, clients cannot reason abstractly, i.e. given only the program text, about types and binding. Instead, they must reason transparently, i.e. about the underlying expansion. To address these problems, this paper  formally introduces \emph{simple} and \emph{parametric typed literal macros} (TLMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning that 1) expansions can refer to external bindings only via spliced terms or supplied module parameters; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
TLMs offer substantially more syntactic control than existing hygienic term-rewriting macro systems while ensuring that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced terms. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We specify TLMs and establish these abstract reasoning principles formally using an ML-style calculus of expressions, patterns and modules.

% \emph{Typed syntax macros (TLMs)}, proposed in a recent short paper by Omar et al. \cite{sac15}, give library providers programmatic control over the parsing and expansion of only terms of {(generalized) literal form}. This appears to occupy a ``sweet spot'' in that it captures many common syntactic idioms while avoiding the problem of conflict. TLMs also maintain a reasonable type and binding discipline.% In particular, clients can use any combination of TLMs in a program without needing to consider conflicts between them, and the language validates each expansion that a TLM generates to maintain 1) a \emph{type discipline} (meaning clients can determine the type of an unexpanded expression without examining its expansion directly); 
% %and 2) a \emph{hygienic binding discipline}.
% %, meaning that the expansion cannot make any assumptions about bindings at the application site, nor  introduce ``hidden bindings'' into subterms.

@@

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
When defining the textual syntax of a programming language, it is common practice to include ``syntactic sugar'' specific to certain standard library constructs.   
For example, Standard ML and other functional languages include sugar for constructing and pattern matching on lists, as defined in the standard library \cite{mthm97-for-dart}.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
%Many languages define convenient list literal forms. This paper questions this common practice: why should an otherwise ordinary data structure be privileged in the definition of a general-purpose programming language? 
When designing the surface syntax of a programming language, it is common practice to include literal forms for a select few standard data structures, e.g. lists. This practice neglects an unbounded assortment of other data structures for which literal forms might also be useful (consider, for example, sets, maps, matrices, regular expressions, HTML trees, URLs, SQL queries and chemical structures.) A more parsimonious approach would be to allow library providers to define new literal forms in a decentralized manner, but the mechanisms of  available to programmers today, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, are, to put it pithily, \emph{unreasonable}. In particular,  they do not support modular reasoning about syntactic determinism, so separately defined literal forms can and do conflict syntactically with one another. Moreover, clients cannot reason abstractly about types and binding when they encounter an unfamiliar literal form (like they can when they encounter an unfamiliar function being applied.) Instead, they must reason transparently, i.e. about the underlying expansion. This increases cognitive cost, and therefore defeats much of the purpose of syntactic sugar. %This makes it difficult to program ``in the large'', and these mechanisms have not been widely adopted.% Hygienic, typed term-rewriting systems, e.g. the Scala macro system \cite{ScalaMacros2013}, are only somewhat more reasonable, and do not offer direct syntactic control.

This paper formally details a more reasonable alternative: \emph{typed literal macros} (TLMs). TLMs give library providers full programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning 1) that expansions cannot make any assumptions about the application site context; and 2) that expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced segments. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We establish these abstract reasoning principles formally with a calculus of expressions, patterns and ML-style modules.  This is the first detailed type-theoretic account of a typed, hygienic term-rewriting macro system for an ML-like language. % This calculus is the first such formalization of a typed macro system.

% \emph{Typed syntax matcros (TLMs)}, proposed in a recent short paper by Omar et al. \cite{sac15}, give library providers programmatic control over the parsing and expansion of only terms of {(generalized) literal form}. This appears to occupy a ``sweet spot'' in that it captures many common syntactic idioms while avoiding the problem of conflict. TLMs also maintain a reasonable type and binding discipline.% In particular, clients can use any combination of TLMs in a program without needing to consider conflicts between them, and the language validates each expansion that a TLM generates to maintain 1) a \emph{type discipline} (meaning clients can determine the type of an unexpanded expression without examining its expansion directly); 
% %and 2) a \emph{hygienic binding discipline}.
% %, meaning that the expansion cannot make any assumptions about bindings at the application site, nor  introduce ``hidden bindings'' into subterms.

@@

\newcommand{\introSec}{Introduction}
\section{\protect\introSec}
\label{sec:intro}
When defining the surface syntax of a programming language, it is common practice to include a few ``literal forms'' specific to certain standard data structures.   
For example, Standard ML and other functional languages include sugar for constructing and pattern matching on lists, as defined in the standard library \cite{mthm97-for-dart}. For example, programmers can write \li{[x, y, z]} rather than \li{Cons(x, Cons(y, Cons(z, Nil)))}.
